Nock for Everyday Coders

### 2025-07-03 mss.tex
Make code blocks more consistent.

\begin{lstlisting}[style=listingcode]
/[(a + a) b]        /[2 /[a b]]
\end{lstlisting}

@@

\begin{lstlisting}[style=listingcode]
[6
 [5 [0 6] [1 0]]
 [0 7]
 [9 4 [[0 2] [2 [0 6] [0 5]] [4 0 7]]]
]
\end{lstlisting}

@@

\begin{lstlisting}[style=listingcode]
:: a number
782
:: a cell (pair with two elements)
[782 9872]
:: each element can itself be a pair
[782 [9872 89728]]
:: the above can be written as
[782 9872 89728]
\end{lstlisting}

@@

\begin{enumerate}
  \item  Start up a Dojo session on a fake ship.\footnote{Consult \texttt{docs.urbit.org} for details.}
  \item  At the prompt, we can execute Nock using \texttt{.*} dottar, e.g., \lstinline[style=inlinecode]{.*(NOCK_SUBJECT, NOCK_FORMULA)}.
\end{enumerate}

@@

\begin{lstlisting}[style=listingcode]
> .*(42 [0 1])
42
\end{lstlisting}

@@

Noun tree addressing is directly addressed in the Nock specification:

@@

\noindent{}
This permits the address to be defined within a particular subtree as well as within the overall tree (Figure \ref{fig:treefull}).

@@

\begin{lstlisting}[style=listingcode]
[[4 5] [6 14 15]]
\end{lstlisting}

@@

\begin{lstlisting}[style=listingcode]
> .*([20 30] [@1@ 67])
67
:: [@1@ 2 587] is same as [@1@ [2 [587]]]
> .*([20 30] [@1@ [2 [587])
[2 587]
\end{lstlisting}

@@

\begin{lstlisting}[style=listingcode]
> .*(50 [@4@ @0@ 1])
51

@@

\begin{lstlisting}[style=listingcode]
> .*(50 [@4@ @0@ 1])
51
:: change to pseudocode
*[50 [@4@ @0@ 1]]
:: move the 4 to the outside as +
+*[50 [@0@ 1]]
:: expand the 0 opcode to the memory slot operator "/"
+/[@1@ 50]
:: grabs the memory slot
+(50)
:: evaluate
51
\end{lstlisting}

@@

\begin{lstlisting}[style=listingcode]
[50 [@0@ 1]]
\end{lstlisting}

@@

\begin{lstlisting}[style=listingcode]
> .*(50 [@4@ @4@ @0@ 1])
52
*[50 [@4@ @4@ @0@ 1]]
:: the first 4 moves outside as a '+'
+*[50 [@4@ @0@ 1]]
:: 2nd 4 becomes a '+'...we have opcode 0 again!
++*[50 [@0@ 1]]
++/[@1@ 50]
++(50)
+(51)
52
\end{lstlisting}

@@

\begin{lstlisting}[style=listingcode]
> .*([100 150] [@4@ @4@ @0@ 3])
152
a (the subject) = [100 150]
*[[100 150] [@4@ @4@ @0@ 3]]
+*[[100 150] [@4@ @0@ 3]]
:: Now we've extracted all the increments,
:: so we just grab the value at memory slot 3
++*[[100 150] [@0@ 3]]
++/[3 [100 150]]
++(150)
+(151)
152
\end{lstlisting}

@@

\begin{lstlisting}[style=listingcode]
> .*(50 [@4@ @1@ 98])
99
*[50 [@4@ @1@ 98]]
:: formula is the "quoter" function
+*[50 [@1@ 98]]
+(98)
99
\end{lstlisting}

@@

\begin{lstlisting}[style=listingcode]
> .*(50 [@4@ @1@ [0 2]])
dojo: hoon expression failed
*[50 [@4@ @1@ [0 2]]]
:: OK cool, the nested value is a 1 opcode
+*[50 [@1@ [0 2]]]
:: 1 ignores the subject (50) and just returns [0 2]
+([0 2])
:: [0 2] isn't an atom, so how can we increment it???
:: We can't, so we crash.
dojo: hoon expression failed
\end{lstlisting}

@@

\begin{lstlisting}[style=listingcode]
> .*(50 [@0@ 1])
50
\end{lstlisting}

@@

\begin{lstlisting}[style=listingcode]
> .*(50 0)
dojo: hoon expression failed
\end{lstlisting}

@@

\begin{lstlisting}[style=listingcode]
> .*(50 [[@0@ 1] [@1@ 203]])
[50 203]
\end{lstlisting}

@@

\begin{lstlisting}[style=listingcode]
> .*(50 [[@0@ 1] [@1@ 203] [@0@ 1] [@1@ 19] [@1@ 76]])
[50 203 50 19 76]
\end{lstlisting}

@@

\begin{lstlisting}[style=listingcode]
> .*([19 20] [[@0@ 1] [@1@ 76] [@4@ @4@ @0@ 3]])
[[19 20] 76 22]
\end{lstlisting}

@@

\begin{lstlisting}[style=listingcode]
[@8@ [@1@ 0] @8@ [@1@ 6 [@5@ [@0@ 7] @4@ @0@ 6] [@0@ 6] @9@ 2 [@0@ 2]
            [@4@ @0@ 6] @0@ 7] @9@ 2 @0@ 1]
\end{lstlisting}

@@

\begin{lstlisting}[style=listingcode]
> .*(100 [@8@ [@1@ 0] @8@ [@1@ 6 [@5@ [@0@ 7] @4@ @0@ 6] [@0@ 6] @9@ 2 [@0@ 2] [@4@ @0@ 6] @0@ 7] @9@ 2 @0@ 1])
99
\end{lstlisting}

@@

\begin{lstlisting}[style=listingcode]
> .*([50 51] some-formula)
\end{lstlisting}

@@

The function/opcode \lstinline[style=inlinecode]{@2@} is designd to handle this problem for us. First the pseudocode:

\begin{lstlisting}[style=listingcode]
:: PSEUDOCODE
*[a @2@ b c]  *[*[a b] *[a c]]
\end{lstlisting}

@@

\begin{lstlisting}[style=listingcode]
*[*[a b] *[a c]]
\end{lstlisting}

@@

\begin{lstlisting}[style=listingcode]
*[a @4@ @4@ b]
:: a: 59
:: b: [1 0]
*[59 @4@ @4@ [@1@ 0]]
:: substitute out the two increment operators
++*[59 [1 0]]
:: ignore subject, return quoted value 0
++(0)
2
\end{lstlisting}

@@

\begin{lstlisting}[style=listingcode]
:: remember, "result-of-step-4" was 2
*[[2 3] @0@ result-of-step-4]
*[[2 3] @0@ 2]
:: get memory slot 2
2
\end{lstlisting}

@@

\begin{lstlisting}[style=listingcode]
*[[2 3] @0@ 10]
:: there's no slot 10
CRASH!!
\end{lstlisting}

@@

\begin{lstlisting}[style=listingcode]
:: if step 3 returned "2" (true)
*[[[@0@ 1] [@1@ 203]] @0@ 2]
[@0@ 1]
\end{lstlisting}

@@

\begin{lstlisting}[style=listingcode]
*[a formula-from-step-2]
:: let's say we returned formula [0 1]
:: our original a, from step 4, was 59
*[59 [@0@ 1]]
59
\end{lstlisting}

@@

\noindent{}
... but in English, this is just saying:

@@

\noindent{}
The fact that the above description has the words ``make a new subject" tells us right away that it's syntactic sugar for opcode \lstinline[style=inlinecode]{@2@}, and that's indeed what we see in the pseudocode.

@@

\begin{lstlisting}[style=listingcode]
a: 45
b: 2
c: [[@1@ @4@ @0@ 3] @0@ 1]
\end{lstlisting}

@@

\begin{lstlisting}[style=listingcode]
[[@1@ @4@ @0@ 3] [@0@ 1]]
\end{lstlisting}

@@

\noindent{}
You can think of this as setting up a subject, pulling a stored procedure from it, and then running that procedure against the subject.

@@

\noindent{}
It replaces the memory slot \lstinline[style=inlinecode]{mem-slot} in \lstinline[style=inlinecode]{target-tree} with \lstinline[style=inlinecode]{new-val}.

\begin{lstlisting}[style=listingcode]
::  Example
#[2 [4 5] [99 88 77]]
[[4 5] 88 77]
\end{lstlisting}

\noindent{}
The pseudocode for \lstinline[style=inlinecode]{@10@} is:

@@

\noindent{}
So we:

@@

\noindent{}
We know the below code is a gate that evaluates its sample, and returns it if it's a loobean, and crashes otherwise. Let's see how that works.

\begin{lstlisting}[style=listingcode]
[@6@ [@5@ [@1@ 0] @0@ 6] [@1@ 0] @6@ [@5@ [@1@ 1] @0@ 6] [@1@ 1] @0@ 0]
\end{lstlisting}

\noindent{}
We start with \lstinline[style=inlinecode]{@6@}, which means this is an if-else. The true/false test is the next element:

\begin{lstlisting}[style=listingcode]
[@5@ [@1@ 0] @0@ 6]
\end{lstlisting}

This compares the quoted value \lstinline[style=inlinecode]{0} (from \lstinline[style=inlinecode]{[@1@ 0]}) with the value at memory slot \lstinline[style=inlinecode]{6} in the subject (\lstinline[style=inlinecode]{[@0@ 6]}).

@@

\begin{lstlisting}[style=listingcode]
[@6@ [@5@ [@1@ 1] @0@ 6] [@1@ 1] @0@ 0]
\end{lstlisting}

@@

If not, it means our input was neither \lstinline[style=inlinecode]{0} nor \lstinline[style=inlinecode]{1} and is not a loobean, so we run the formula \lstinline[style=inlinecode]{[@0@ 0]}, which always crashes (as there is no memory slot \lstinline[style=inlinecode]{0}).\footnote{In the binary tree addressing scheme of Nock, 1 refers to the entire noun, 2 to its head if it exists, and 3 to its tail if it exists.} This is exactly what we want—crash if the sample is not a loobean!

@@

\begin{lstlisting}[style=listingcode]
*[50 @4@ @4@ [@5@ [@0@ 1] [@1@ 50]]]
\end{lstlisting}

\noindent{}
The interpreter can then be thought of as moving left-to-right, expanding according to its rules, until it can't expand any further into pseudocode:

\begin{lstlisting}[style=listingcode]
*[50 @4@ @4@ [@5@ [@0@ 1] [@1@ 50]]]
+*[50 @4@ [@5@ [@0@ 1] [@1@ 50]]]
++*[50 @5@ [@0@ 1] [@1@ 50]]
++=[*[50 [@0@ 1]] *[50 [@1@ 50]]]
++=[/[@1@ 50] 50]
\end{lstlisting}

\noindent{}
Once we get to that last line, \lstinline[style=inlinecode]{++=[/[@1@ 50] *[50 [@1@ 50]]]}, no further expansion into pseudocode is possible.

@@

\begin{lstlisting}[style=listingcode]
++=[50 50]
\end{lstlisting}

\noindent{}
Then it moves to the next-furthest-inside operator: \lstinline[style=inlinecode]{=}, and then to the \lstinline[style=inlinecode]{+} operators:

\begin{lstlisting}[style=listingcode]
++=[50 50]
++(0)
+(1)
2
\end{lstlisting}

@@

\begin{lstlisting}[style=listingcode]
::  head is an atom (b)
*[a @11@ b c]  *[a c]

@@

\begin{lstlisting}[style=listingcode]
.*([50 51] [@11@ 369 @0@ 2])
\end{lstlisting}

\noindent{}
This passes the value \lstinline[style=inlinecode]{369} to the interpreter, where maybe it would be the key in a hashmap. The value associated with the could be the function to debug print the whole operation. Or the value could be a jet function that specifically knows how to compute the \lstinline[style=inlinecode]{@0@} opcode faster when it's followed by a \lstinline[style=inlinecode]{@2@}.

@@

\begin{lstlisting}[style=listingcode]
> .*([50 51] [@0@ 2])
50
\end{lstlisting}

@@

\begin{lstlisting}[style=listingcode]
> =increment-formula [@4@ @0@ 1]
> .*(50 increment-formula)
51
> =increment-formula
\end{lstlisting}

@@

\begin{lstlisting}[style=listingcode]
> =inc-gate [inc 0]
> inc-gate
[[@7@ [@0@ 3] @4@ @0@ 1] 0]
\end{lstlisting}

@@

\begin{lstlisting}[style=listingcode]
> .*(inc-gate [@9@ 2 @0@ 1])
1
\end{lstlisting}

@@

\begin{lstlisting}[style=listingcode]
:: notice how opcode 10 can replace the last element of inc-gate
:: the tail of inc-gate was 0; now it's 36
> .*([inc-gate 36] [@10@ [@3@ [@0@ 3]] @0@ 2])
[[@7@ [@0@ 3] @4@ @0@ 1] 36]

@@

\begin{lstlisting}[style=listingcode]
::  gets value in mem slot 3 and applies the decrement formula to it
> =dec-arm [@7@ [@0@ 3] dec]

@@

\begin{lstlisting}[style=listingcode]
:: all we need to do is add the sample! (we give it a default value of 0)
> =dec-gate [dec-arm 0]
> dec-gate
[[7 [0 3] 6 [5 [0 1] 1 0] [0 0] 6 [3 0 1] [0 0] 8 [1 0] 8 [1 6 [5 [0 7] 4 0 6] [0 6] 9 2 [0 2] [4 0 6] 0 7] 9 2 0 1] 0]
[[@7@ [@0@ 3] @6@ [@5@ [@0@ 1] @1@ 0] [@0@ 0] @6@ [@3@ @0@ 1] [@0@ 0] @8@ [@1@ 0] @8@ [@1@ 6 [@5@ [@0@ 7] @4@ @0@ 6] [@0@ 6] @9@ 2 [@0@ 2] [@4@ @0@ 6] @0@ 7] @9@ 2 @0@ 1] 0]
\end{lstlisting}

@@

\begin{lstlisting}[style=listingcode]
::  confirm that our 10 code places 36 in the tail
> .*([dec-gate 36] [@10@ [@3@ [@0@ 3]] @0@ 2])
[[@7@ [@0@ 3] @6@ [@5@ [@0@ 1] @1@ 0] [@0@ 0] @6@ [@3@ @0@ 1] [@0@ 0] @8@ [@1@ 0] @8@ [@1@ 6 [@5@ [@0@ 7] @4@ @0@ 6] [@0@ 6] @9@ 2 [@0@ 2] [@4@ @0@ 6] @0@ 7] @9@ 2 @0@ 1] 36]

@@

\begin{lstlisting}[style=listingcode]
[[main-logic dec-gate] a b]
\end{lstlisting}

@@

\begin{lstlisting}[style=listingcode]
> =dec-a [@9@ 2 @10@ [@3@ [@0@ 6]] @0@ 5]
> =dec-b [@9@ 2 @10@ [@3@ [@0@ 7]] @0@ 5]

@@

\begin{lstlisting}[style=listingcode]
> =recur [@9@ 4 [[@0@ 2] dec-a dec-b]]

@@

\begin{lstlisting}[style=listingcode]
> =battery [main-logic dec-gate]

@@

\begin{lstlisting}[style=listingcode]
[@6@ [@5@ [@0@ 2] @0@ 3] [@1@ 0] @9@ 4 [@1@ [@6@ [@5@ [@0@ 6] @1@ 0] [@1@ 2] @6@ [@5@ [@0@ 7] @1@ 0] [@1@ 1] @9@ 4 [@0@ 2] [@9@ 2 @10@ [@3@ @0@ 6] @0@ 5] @9@ 2 @10@ [@3@ @0@ 7] @0@ 5] [@7@ [@0@ 3] @6@ [@5@ [@0@ 1] @1@ 0] [@0@ 0] @6@ [@3@ @0@ 1] [@0@ 0] @8@ [@1@ 0] @8@ [@1@ @6@ [@5@ [@0@ 7] 4 @0@ 6] [@0@ 6] @9@ 2 [@0@ 2] [@4@ @0@ 6] @0@ 7] @9@ 2 @0@ 1] 0] [@0@ 2] @0@ 3]
\end{lstlisting}

>>>>

\begin{lstlisting}[style=listingblock]
/[(a + a) b]        /[2 /[a b]]
\end{lstlisting}

@@

\begin{lstlisting}[style=listingblock]
[6
 [5 [0 6] [1 0]]
 [0 7]
 [9 4 [[0 2] [2 [0 6] [0 5]] [4 0 7]]]
]
\end{lstlisting}

@@

\begin{lstlisting}[style=listingblock]
:: a number
782
:: a cell (pair with two elements)
[782 9872]
:: each element can itself be a pair
[782 [9872 89728]]
:: the above can be written as
[782 9872 89728]
\end{lstlisting}

@@

\begin{enumerate}
  \item  Start up a Dojo session on a fake ship.\footnote{Consult \url{https://docs.urbit.org} for details.}
  \item  At the prompt, we can execute Nock using \texttt{.*} dottar, e.g., \lstinline[style=inlinecode]{.*(NOCK_SUBJECT, NOCK_FORMULA)}.
\end{enumerate}

@@

\begin{lstlisting}[style=listingblock]
> .*(42 [0 1])
42
\end{lstlisting}

@@

The Nock specification defines noun tree addressing:

@@

\noindent
This permits the address to be defined within a particular subtree as well as within the overall tree (Figure \ref{fig:treefull}).

@@

\begin{lstlisting}[style=listingblock]
[[4 5] [6 14 15]]
\end{lstlisting}

@@

\begin{lstlisting}[style=listingblock]
> .*([20 30] [@1@ 67])
67
:: [@1@ 2 587] is same as [@1@ [2 [587]]]
> .*([20 30] [@1@ [2 [587])
[2 587]
\end{lstlisting}

@@

\begin{lstlisting}[style=listingblock]
> .*(50 [@4@ @0@ 1])
51

@@

\begin{lstlisting}[style=listingblock]
> .*(50 [@4@ @0@ 1])
51
:: change to pseudocode
*[50 [@4@ @0@ 1]]
:: move the 4 to the outside as +
+*[50 [@0@ 1]]
:: expand the 0 opcode to the memory slot operator "/"
+/[@1@ 50]
:: grabs the memory slot
+(50)
:: evaluate
51
\end{lstlisting}

@@

\begin{lstlisting}[style=listingblock]
[50 [@0@ 1]]
\end{lstlisting}

@@

\begin{lstlisting}[style=listingblock]
> .*(50 [@4@ @4@ @0@ 1])
52
*[50 [@4@ @4@ @0@ 1]]
:: the first 4 moves outside as a '+'
+*[50 [@4@ @0@ 1]]
:: 2nd 4 becomes a '+'...we have opcode 0 again!
++*[50 [@0@ 1]]
++/[@1@ 50]
++(50)
+(51)
52
\end{lstlisting}

@@

\begin{lstlisting}[style=listingblock]
> .*([100 150] [@4@ @4@ @0@ 3])
152
a (the subject) = [100 150]
*[[100 150] [@4@ @4@ @0@ 3]]
+*[[100 150] [@4@ @0@ 3]]
:: Now we've extracted all the increments,
:: so we just grab the value at memory slot 3
++*[[100 150] [@0@ 3]]
++/[3 [100 150]]
++(150)
+(151)
152
\end{lstlisting}

@@

\begin{lstlisting}[style=listingblock]
> .*(50 [@4@ @1@ 98])
99
*[50 [@4@ @1@ 98]]
:: formula is the "quoter" function
+*[50 [@1@ 98]]
+(98)
99
\end{lstlisting}

@@

\begin{lstlisting}[style=listingblock]
> .*(50 [@4@ @1@ [0 2]])
dojo: hoon expression failed
*[50 [@4@ @1@ [0 2]]]
:: OK cool, the nested value is a 1 opcode
+*[50 [@1@ [0 2]]]
:: 1 ignores the subject (50) and just returns [0 2]
+([0 2])
:: [0 2] isn't an atom, so how can we increment it???
:: We can't, so we crash.
dojo: hoon expression failed
\end{lstlisting}

@@

\begin{lstlisting}[style=listingblock]
> .*(50 [@0@ 1])
50
\end{lstlisting}

@@

\begin{lstlisting}[style=listingblock]
> .*(50 0)
dojo: hoon expression failed
\end{lstlisting}

@@

\begin{lstlisting}[style=listingblock]
> .*(50 [[@0@ 1] [@1@ 203]])
[50 203]
\end{lstlisting}

@@

\begin{lstlisting}[style=listingblock]
> .*(50 [[@0@ 1] [@1@ 203] [@0@ 1] [@1@ 19] [@1@ 76]])
[50 203 50 19 76]
\end{lstlisting}

@@

\begin{lstlisting}[style=listingblock]
> .*([19 20] [[@0@ 1] [@1@ 76] [@4@ @4@ @0@ 3]])
[[19 20] 76 22]
\end{lstlisting}

@@

\begin{lstlisting}[style=listingblock]
[@8@ [@1@ 0] @8@ [@1@ 6 [@5@ [@0@ 7] @4@ @0@ 6] [@0@ 6] @9@ 2 [@0@ 2]
            [@4@ @0@ 6] @0@ 7] @9@ 2 @0@ 1]
\end{lstlisting}

@@

\begin{lstlisting}[style=listingblock]
> .*(100 [@8@ [@1@ 0] @8@ [@1@ 6 [@5@ [@0@ 7] @4@ @0@ 6] [@0@ 6] @9@ 2 [@0@ 2] [@4@ @0@ 6] @0@ 7] @9@ 2 @0@ 1])
99
\end{lstlisting}

@@

\begin{lstlisting}[style=listingblock]
> .*([50 51] some-formula)
\end{lstlisting}

@@

The function/opcode \lstinline[style=inlinecode]{@2@} is designed to handle this problem for us. First the pseudocode:

\begin{lstlisting}[style=listingcode]
*[a @2@ b c]  *[*[a b] *[a c]]
\end{lstlisting}

@@

\begin{lstlisting}[style=listingblock]
*[*[a b] *[a c]]
\end{lstlisting}

@@

\begin{lstlisting}[style=listingblock]
*[a @4@ @4@ b]
:: a: 59
:: b: [1 0]
*[59 @4@ @4@ [@1@ 0]]
:: substitute out the two increment operators
++*[59 [1 0]]
:: ignore subject, return quoted value 0
++(0)
2
\end{lstlisting}

@@

\begin{lstlisting}[style=listingblock]
:: remember, "result-of-step-4" was 2
*[[2 3] @0@ result-of-step-4]
*[[2 3] @0@ 2]
:: get memory slot 2
2
\end{lstlisting}

@@

\begin{lstlisting}[style=listingblock]
*[[2 3] @0@ 10]
:: there's no slot 10
CRASH!!
\end{lstlisting}

@@

\begin{lstlisting}[style=listingblock]
:: if step 3 returned "2" (true)
*[[[@0@ 1] [@1@ 203]] @0@ 2]
[@0@ 1]
\end{lstlisting}

@@

\begin{lstlisting}[style=listingblock]
*[a formula-from-step-2]
:: let's say we returned formula [0 1]
:: our original a, from step 4, was 59
*[59 [@0@ 1]]
59
\end{lstlisting}

@@

\noindent
... but in English, this is just saying:

@@

\noindent
The fact that the above description has the words ``make a new subject" tells us right away that it's syntactic sugar for opcode \lstinline[style=inlinecode]{@2@}, and that's indeed what we see in the pseudocode.

@@

\begin{lstlisting}[style=listingblock]
a: 45
b: 2
c: [[@1@ @4@ @0@ 3] @0@ 1]
\end{lstlisting}

@@

\begin{lstlisting}[style=listingblock]
[[@1@ @4@ @0@ 3] [@0@ 1]]
\end{lstlisting}

@@

\noindent
You can think of this as setting up a subject, pulling a stored procedure from it, and then running that procedure against the subject.

@@

\noindent
It replaces the memory slot \lstinline[style=inlinecode]{mem-slot} in \lstinline[style=inlinecode]{target-tree} with \lstinline[style=inlinecode]{new-val}.

\begin{lstlisting}[style=listingblock]
::  Example
#[2 [4 5] [99 88 77]]
[[4 5] 88 77]
\end{lstlisting}

\noindent
The pseudocode for \lstinline[style=inlinecode]{@10@} is:

@@

\noindent
So we:

@@

\noindent
The below code acts as a formula which evaluates its sample, and returns it if it's a loobean or crashes otherwise. Let's see how that works.

\begin{lstlisting}[style=listingblock]
[@6@ [@5@ [@1@ 0] @0@ 6] [@1@ 0] @6@ [@5@ [@1@ 1] @0@ 6] [@1@ 1] @0@ 0]
\end{lstlisting}

\noindent
We start with \lstinline[style=inlinecode]{@6@}, which means this is an if-else. The true/false test is the next element:

\begin{lstlisting}[style=listingblock]
[@5@ [@1@ 0] @0@ 6]
\end{lstlisting}

\noindent
This compares the quoted value \lstinline[style=inlinecode]{0} (from \lstinline[style=inlinecode]{[@1@ 0]}) with the value at memory slot \lstinline[style=inlinecode]{6} in the subject (\lstinline[style=inlinecode]{[@0@ 6]}).

@@

\begin{lstlisting}[style=listingblock]
[@6@ [@5@ [@1@ 1] @0@ 6] [@1@ 1] @0@ 0]
\end{lstlisting}

@@

If not, it means our input was neither \lstinline[style=inlinecode]{0} nor \lstinline[style=inlinecode]{1} and is not a loobean, so we run the formula \lstinline[style=inlinecode]{[@0@ 0]}, which always crashes (as there is no memory slot \lstinline[style=inlinecode]{0}).\footnote{Recall that in the binary tree addressing scheme of Nock, 1 refers to the entire noun, 2 to its head if it exists, and 3 to its tail if it exists.} This is exactly what we want—crash if the sample is not a loobean!

The second part of the formula which is composed with an \lstinline[style=inlinecode]{@8@} is:

\begin{lstlisting}[style=listingblock]
[@8@ [@5@ [@0@ 14] @0@ 2] @0@ 6]
\end{lstlisting}

This checks whether the battery of the current gate is equivalent to the head of the context; if so, it returns the sample.  This is a artifact of a type promotion in Hoon which converts a bare loobean value to a constant by pinning a constant as an example first (i.e., at slot 14).  (It's completely extraneous in Nock itself, however.)

@@

\begin{lstlisting}[style=listingblock]
*[50 @4@ @4@ [@5@ [@0@ 1] [@1@ 50]]]
\end{lstlisting}

\noindent
The interpreter can then be thought of as moving left-to-right, expanding according to its rules, until it can't expand any further into pseudocode:

\begin{lstlisting}[style=listingblock]
*[50 @4@ @4@ [@5@ [@0@ 1] [@1@ 50]]]
+*[50 @4@ [@5@ [@0@ 1] [@1@ 50]]]
++*[50 @5@ [@0@ 1] [@1@ 50]]
++=[*[50 [@0@ 1]] *[50 [@1@ 50]]]
++=[/[@1@ 50] 50]
\end{lstlisting}

\noindent
Once we get to that last line, \lstinline[style=inlinecode]{++=[/[@1@ 50] *[50 [@1@ 50]]]}, no further expansion into pseudocode is possible.

@@

\begin{lstlisting}[style=listingblock]
++=[50 50]
\end{lstlisting}

\noindent
Then it moves to the next-furthest-inside operator: \lstinline[style=inlinecode]{=}, and then to the \lstinline[style=inlinecode]{+} operators:

\begin{lstlisting}[style=listingblock]
++=[50 50]
++(0)
+(1)
2
\end{lstlisting}

@@

\begin{lstlisting}[style=listingblock]
::  head is an atom (b)
*[a @11@ b c]  *[a c]

@@

\begin{lstlisting}[style=listingblock]
.*([50 51] [@11@ 369 @0@ 2])
\end{lstlisting}

\noindent
This passes the value \lstinline[style=inlinecode]{369} to the interpreter, where maybe it would be the key in a hashmap. The value associated with the could be the function to debug print the whole operation. Or the value could be a jet function that specifically knows how to compute the \lstinline[style=inlinecode]{@0@} opcode faster when it's followed by a \lstinline[style=inlinecode]{@2@}.

@@

\begin{lstlisting}[style=listingblock]
> .*([50 51] [@0@ 2])
50
\end{lstlisting}

@@

\begin{lstlisting}[style=listingblock]
> =increment-formula [@4@ @0@ 1]
> .*(50 increment-formula)
51
> =increment-formula
\end{lstlisting}

@@

\begin{lstlisting}[style=listingblock]
> =inc-gate [inc 0]
> inc-gate
[[@7@ [@0@ 3] @4@ @0@ 1] 0]
\end{lstlisting}

@@

\begin{lstlisting}[style=listingblock]
> .*(inc-gate [@9@ 2 @0@ 1])
1
\end{lstlisting}

@@

\begin{lstlisting}[style=listingblock]
:: notice how opcode 10 can replace the last element of inc-gate
:: the tail of inc-gate was 0; now it's 36
> .*([inc-gate 36] [@10@ [@3@ [@0@ 3]] @0@ 2])
[[@7@ [@0@ 3] @4@ @0@ 1] 36]

@@

\begin{lstlisting}[style=listingblock]
::  gets value in mem slot 3 and applies the decrement formula to it
> =dec-arm [@7@ [@0@ 3] dec]

@@

\begin{lstlisting}[style=listingblock]
:: all we need to do is add the sample! (we give it a default value of 0)
> =dec-gate [dec-arm 0]
> dec-gate
[[7 [0 3] 6 [5 [0 1] 1 0] [0 0] 6 [3 0 1] [0 0] 8 [1 0] 8 [1 6 [5 [0 7] 4 0 6] [0 6] 9 2 [0 2] [4 0 6] 0 7] 9 2 0 1] 0]
[[@7@ [@0@ 3] @6@ [@5@ [@0@ 1] @1@ 0] [@0@ 0] @6@ [@3@ @0@ 1] [@0@ 0] @8@ [@1@ 0] @8@ [@1@ 6 [@5@ [@0@ 7] @4@ @0@ 6] [@0@ 6] @9@ 2 [@0@ 2] [@4@ @0@ 6] @0@ 7] @9@ 2 @0@ 1] 0]
\end{lstlisting}

@@

\begin{lstlisting}[style=listingblock]
::  confirm that our 10 code places 36 in the tail
> .*([dec-gate 36] [@10@ [@3@ [@0@ 3]] @0@ 2])
[[@7@ [@0@ 3] @6@ [@5@ [@0@ 1] @1@ 0] [@0@ 0] @6@ [@3@ @0@ 1] [@0@ 0] @8@ [@1@ 0] @8@ [@1@ 6 [@5@ [@0@ 7] @4@ @0@ 6] [@0@ 6] @9@ 2 [@0@ 2] [@4@ @0@ 6] @0@ 7] @9@ 2 @0@ 1] 36]

@@

\begin{lstlisting}[style=listingblock]
[[main-logic dec-gate] a b]
\end{lstlisting}

@@

\begin{lstlisting}[style=listingblock]
> =dec-a [@9@ 2 @10@ [@3@ [@0@ 6]] @0@ 5]
> =dec-b [@9@ 2 @10@ [@3@ [@0@ 7]] @0@ 5]

@@

\begin{lstlisting}[style=listingblock]
> =recur [@9@ 4 [[@0@ 2] dec-a dec-b]]

@@

\begin{lstlisting}[style=listingblock]
> =battery [main-logic dec-gate]

@@

\begin{lstlisting}[style=listingblock]
[@6@ [@5@ [@0@ 2] @0@ 3] [@1@ 0] @9@ 4 [@1@ [@6@ [@5@ [@0@ 6] @1@ 0] [@1@ 2] @6@ [@5@ [@0@ 7] @1@ 0] [@1@ 1] @9@ 4 [@0@ 2] [@9@ 2 @10@ [@3@ @0@ 6] @0@ 5] @9@ 2 @10@ [@3@ @0@ 7] @0@ 5] [@7@ [@0@ 3] @6@ [@5@ [@0@ 1] @1@ 0] [@0@ 0] @6@ [@3@ @0@ 1] [@0@ 0] @8@ [@1@ 0] @8@ [@1@ @6@ [@5@ [@0@ 7] 4 @0@ 6] [@0@ 6] @9@ 2 [@0@ 2] [@4@ @0@ 6] @0@ 7] @9@ 2 @0@ 1] 0] [@0@ 2] @0@ 3]
\end{lstlisting}
